---
title: "Bonus Tools"
format: html
engine: knitr
vignette: >
  %\VignetteIndexEntry{Bonus estimators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This short vignette collects examples for the "bonus" model families in
`dbreg`: `dbbinsreg`, `dbkreg`, `dbdml`, and `dbforest`.

Here, "bonus" means estimators that sit outside the core `dbreg` linear-model
workflow, but are layered on top of the core functionality therein and follow the same database-first design.
They are optional tools for specific tasks: binscatter diagnostics, nonparametric
smoothing, causal partially linear estimation, and tree-based prediction.

```{r}
library(dbreg)
set.seed(123)

n = 800
g = factor(sample(1:20, n, replace = TRUE))
x = runif(n, -2, 2)
d = 0.6 * x + as.numeric(g) / 10 + rnorm(n)
y = 1 + 1.8 * d + sin(x) + as.numeric(g) / 5 + rnorm(n)

df = data.frame(y = y, d = d, x = x, g = g)
```

## `dbbinsreg`

`dbbinsreg` is a SQL-first binscatter estimator with optional smooth lines,
pointwise confidence intervals, and confidence bands. The main controls are
`nbins`/`binspos` (partitioning), `points` and `line` (polynomial/smoothness),
and `ci`/`cb` for inference layers.

```{r}
mod_bins = dbbinsreg(
  y ~ x,
  data = df,
  nbins = 20,
  binspos = "qs",
  points = c(1, 1),
  line = c(1, 1),
  plot = FALSE
)
mod_bins
plot(mod_bins, theme = "clean", main = "dbbinsreg: bins + smooth line")
```

## `dbkreg`

`dbkreg` fits kernel smoothers on database backends. Key arguments are
`bandwidth`, `kernel`, and `degree` (`0` local constant, `1` local linear),
plus grid controls (`eval`, `n_eval`, `grid_method`) and optional
pointwise CIs via `ci = TRUE`.

```{r}
mod_kreg = dbkreg(
  y ~ x,
  data = df,
  degree = 1,
  bandwidth = 0.4,
  n_eval = 120
)
mod_kreg
plot(mod_kreg, theme = "clean", main = "dbkreg: local linear fit")
```

```{r}
mod_kreg_ci = dbkreg(
  y ~ x,
  data = df,
  degree = 1,
  bandwidth = 0.4,
  n_eval = 120,
  ci = TRUE,
  level = 0.95
)

gci = mod_kreg_ci$grid[order(mod_kreg_ci$grid$x), ]
tinyplot::tinyplot(
  fit ~ x,
  data = gci,
  type = "n",
  theme = "clean",
  xlab = "x",
  ylab = "y",
  main = "dbkreg: local linear fit with 95% CI"
)
tinyplot::tinyplot_add(
  ymin = gci$lwr,
  ymax = gci$upr,
  type = "ribbon",
  lty = 0
)
tinyplot::tinyplot_add(
  data = gci,
  type = "l"
)
```

For multi-dimensional fits, plotting is currently 1D-only. You can still fit
and inspect multi-D results through `mod$grid`.

```{r eval=FALSE}
grid2 = expand.grid(
  x = seq(-2, 2, length.out = 25),
  d = seq(min(df$d), max(df$d), length.out = 25)
)

mod_kreg_2d = dbkreg(
  y ~ x + d,
  data = df,
  eval = "points",
  grid = grid2,
  degree = 1,
  bandwidth = c(0.4, 0.4)
)
head(mod_kreg_2d$grid)
```

## `dbdml`

`dbdml` provides two estimators through one interface. `method = "strata"`
matches the compressed strata leave-one-out approach for discrete controls,
while `method = "plm"` uses cross-fitted partially linear residualization.
Main arguments are `treat`, `method`, `folds`, and `ridge`.

```{r}
mod_dml_strata = dbdml(
  y ~ d + g,
  data = df,
  treat = "d",
  method = "strata"
)
mod_dml_strata
```

```{r}
mod_dml_plm = dbdml(
  y ~ d + x,
  data = df,
  treat = "d",
  method = "plm",
  folds = 5,
  ridge = 1
)
mod_dml_plm
```

## `dbforest` and `dbtree`

`dbforest` is an ensemble of `dbtree` models. The primary tuning parameters are
`ntree`, `mtry`, `sample_frac`, and `seed`, plus tree-level controls such as
`max_depth`, `min_split`, and `min_leaf`. `dbtree` gives a single interpretable
tree, while `dbforest` targets predictive stability.

```{r}
mod_forest = dbforest(
  y ~ x + d + g,
  data = df,
  ntree = 30,
  mtry = 2,
  sample_frac = 0.8,
  seed = 42,
  max_depth = 4,
  n_bins = 16,
  min_split = 20,
  min_leaf = 8
)
mod_forest
predict(mod_forest, df[1:5, ])
```

```{r}
mod_tree = dbtree(
  y ~ x + d + g,
  data = df,
  max_depth = 3,
  n_bins = 16,
  min_split = 20,
  min_leaf = 8
)
mod_tree
predict(mod_tree, df[1:5, ])
```

To visualize the tree in one dimension, we can fix `d` and `g` and plot
predictions along an `x` slice. This gives an interpretable step profile
for `dbtree` and a smoother ensemble profile for `dbforest`.

```{r echo=FALSE}
g_mode = names(sort(table(df$g), decreasing = TRUE))[1]
d_med = stats::median(df$d)
x_slice = seq(min(df$x), max(df$x), length.out = 200)
slice_df = data.frame(
  x = x_slice,
  d = d_med,
  g = factor(g_mode, levels = levels(df$g))
)
slice_df$tree_pred = as.numeric(predict(mod_tree, slice_df))
slice_df$forest_pred = as.numeric(predict(mod_forest, slice_df))

tinyplot::tinyplot(
  tree_pred ~ x,
  data = slice_df,
  type = "l",
  lwd = 2,
  theme = "clean",
  xlab = "x",
  ylab = "Predicted y",
  main = "1D slice: dbtree vs dbforest"
)
graphics::lines(slice_df$x, slice_df$forest_pred, lwd = 2, lty = 2)
graphics::legend(
  "topleft",
  legend = c("dbtree", "dbforest"),
  lty = c(1, 2),
  lwd = c(2, 2),
  bty = "n"
)
```
